Comparative Analysis of Single--Neuron versus Dual--Neuron Output Layers in Binary Classification Neural Networks

### 2025-05-18 paper/model_architectures.tex
[changes in paper writting [draft]

To provide a comprehensive understanding of the neural network architectures used in our experiments, we present visual diagrams of each model design. These diagrams illustrate the layer configurations, connections, and key architectural features that distinguish each model.

@@

Our custom Small CNN architecture represents a lightweight convolutional neural network designed specifically for this research. Figures~\ref{fig:small_cnn_arch_single} and \ref{fig:small_cnn_arch_dual} illustrate its structure with single-neuron and dual-neuron output configurations, respectively.

@@

The Small CNN architecture features:
\begin{enumerate}
\item Three convolutional blocks, each containing two convolutional layers with batch normalization, followed by max pooling and dropout
\item A fully connected layer with 256 neurons, batch normalization, and dropout
\item Either a single-neuron output with sigmoid activation or a dual-neuron output with softmax activation
\end{enumerate}

This architecture provides a baseline model with approximately 813,000 parameters, all of which are trainable during our experiments.

@@

For our second architecture, we employed the ResNet50 model with transfer learning. Figures~\ref{fig:resnet50_arch_single} and \ref{fig:resnet50_arch_dual} show the architecture with its characteristic residual connections for both output configurations.

@@

Our third architecture is the Vision Transformer (ViT), which represents the attention-based paradigm in computer vision. Figures~\ref{fig:vit_arch_single} and \ref{fig:vit_arch_dual} illustrate the ViT architecture with its distinctive transformer encoder blocks for both output configurations.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.9\textwidth]{figures/vit_1neuron_architecture.png}
\caption{Vision Transformer (ViT) architecture with single-neuron output layer. The model divides the input image into patches, processes them through transformer encoder blocks, and uses the class token representation for classification.}
\label{fig:vit_arch_single}
\end{figure}

@@

The ViT model is the largest in our study with approximately 86 million total parameters, though only about 230,000 of these are trainable due to our transfer learning approach.

@@

As shown in Figures \ref{fig:small_cnn_arch_single}--\ref{fig:vit_arch_dual}, for each architecture, we implemented two output layer configurations:

\begin{enumerate}
\item \textbf{Single-Neuron Configuration}: A single output neuron with sigmoid activation, where outputs closer to 0 represent one class and outputs closer to 1 represent the other class. The loss function used is Binary Cross Entropy.

\item \textbf{Dual-Neuron Configuration}: Two output neurons with softmax activation, where each neuron represents the probability of the input belonging to one of the two classes. The loss function used is Cross Entropy.
\end{enumerate}

As shown in our parameter analysis, the difference in parameter count between these two configurations is minimal (less than 0.1\% increase), ensuring that performance differences can be attributed to the architectural choice rather than model capacity.

>>>>

To provide a detailed understanding of the neural network architectures used in our experiments we show visual diagrams of each model design. These diagrams show the layer configurations, connections, and key architectural features that distinguish each model along with their last output layer neuron.

@@

Our custom Small CNN (developed to generalize our results on custom neural networks if necessary) architecture represents a lightweight convolutional neural network designed specifically for this research. Figures~\ref{fig:small_cnn_arch_single} and \ref{fig:small_cnn_arch_dual} illustrate its structure with single-neuron and dual-neuron output configurations respectively.

@@

The Small CNN architecture features:
\begin{enumerate}
\item Three convolutional blocks each containing two convolutional layers with batch normalization followed by max pooling and dropout
\item A fully connected layer with 256 neurons, batch normalization and dropout
\item Either a single-neuron output with sigmoid activation or a dual-neuron output with softmax activation
\end{enumerate}

This architecture provides a baseline model with approximately 813,000 parameters all of which are trainable during our experiments.

@@

For our second architecture we used the ResNet50 model with transfer learning. Figures~\ref{fig:resnet50_arch_single} and \ref{fig:resnet50_arch_dual} show the architecture with its characteristic residual connections for both output configurations.

@@

Our third architecture is the Vision Transformer (ViT) which represents the attention based paradigm in computer vision. Figures~\ref{fig:vit_arch_single} and \ref{fig:vit_arch_dual} illustrate the ViT architecture with its distinctive transformer encoder blocks for both output configurations.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.9\textwidth]{figures/vit_1neuron_architecture.png}
\caption{Vision Transformer (ViT) architecture with single-neuron output layer. The model divides the input image into patches and processes them through transformer encoder blocks. The class token representation is used for classification.}
\label{fig:vit_arch_single}
\end{figure}

@@

The ViT model is the largest in our study with approximately 86 million total parameters though only about 230,000 of these are trainable due to our transfer learning approach.

@@

As shown in Figures \ref{fig:small_cnn_arch_single}--\ref{fig:vit_arch_dual}, for each architecture we implemented two output layer configurations:

\begin{enumerate}
\item \textbf{Single-Neuron Configuration}: A single output neuron with sigmoid activation where outputs closer to 0 represent one class and outputs closer to 1 represent the other class. The loss function used is Binary Cross Entropy.

\item \textbf{Dual-Neuron Configuration}: Two output neurons with softmax activation where each neuron represents the probability of the input belonging to one of the two classes. The loss function used is Cross Entropy.
\end{enumerate}

As shown in our parameter analysis the difference in parameter count between these two configurations is minimal (less than 0.1\% increase), ensuring that performance differences can be attributed to the architectural choice rather than model capacity.

### 2025-05-18 paper/paper.tex
fixed latex issues
[changes in paper writting [draft]

% Graphics & tables
\usepackage{graphicx}
% \usepackage{booktabs} % Removed for compatibility

@@

\begin{abstract}
Binary classification is a fundamental task in machine learning where models predict one of two possible outcomes. Conventionally, neural networks implement binary classification using a single output neuron with sigmoid activation. However, an alternative approach using two output neurons with softmax activation is also viable. This research provides a comprehensive comparison between these two architectural choices, analyzing their impact on model performance, training dynamics, and generalization capabilities. Through extensive experimentation on standard datasets using multiple neural network architectures, we identify scenarios where each approach may be advantageous. Our findings contribute to the understanding of neural network design considerations for binary classification tasks.
\end{abstract}

\section{Introduction}
In the field of neural networks, binary classification tasks are traditionally approached using a single output neuron with sigmoid activation, where outputs closer to 0 represent one class and outputs closer to 1 represent the other. Despite this convention, an alternative approach exists: using two output neurons with softmax activation, explicitly modeling both the positive and "background" classes. While both approaches can theoretically solve the same problems, their performance characteristics, training dynamics, and generalization capabilities may differ.

This research aims to systematically investigate the differences between these two approaches and their impact on various aspects of model performance. We examine whether the additional expressiveness of a two-neuron output layer provides meaningful advantages over the simpler single-neuron approach, or conversely, whether the more constrained single-neuron architecture offers benefits in terms of regularization and generalization.

@@

\subsection{Neural Network Output Layers for Classification}
Binary classification in neural networks has historically been implemented using a single output neuron with sigmoid activation [4]. The sigmoid function maps the network's output to a value between 0 and 1, which can be interpreted as the probability of the positive class. This approach is mathematically elegant and directly tied to logistic regression.

The alternative approach uses two output neurons with softmax activation, which produces a probability distribution over the two classes [6]. While this is the standard approach for multi-class classification problems, its application to binary problems is less common but theoretically sound.

Variants of the two-neuron softmax design have also been explored for binary problems. Memisevic et al. [6] introduced gated softmax, while Tyagi et al. [5] revisited the sigmoid–versus-softmax debate and showed that a properly tuned sigmoid–MSE objective can outperform softmax–cross-entropy. Further investigations include Yang et al. [13], Klimo et al. [14], Maharjan et al. [15], and Hu et al. [16], which explore alternative loss functions, error-detecting output codes, and hardware-efficient softmax implementations.

\subsection{Related Research}
Prior research on binary image classification can be grouped into three strands: \textbf{output-layer configurations}, \textbf{Vision Transformer (ViT) models}, and \textbf{optimisation \& activation-function studies}.

@@

\textbf{Vision Transformers.} Dosovitskiy et al.'s seminal ViT work [1] demonstrated that pure transformer architectures can match or surpass CNNs on large-scale image classification. Comprehensive surveys [2, 3] and data-efficient variants such as DeiT [11] together with the study by Khalil et al. [12] confirm ViT's competitiveness, motivating its inclusion in our comparison. The transformer paradigm itself originated with Vaswani et al. [20]. Additional surveys and architectural improvements—Khan et al. [17], the hierarchical Swin Transformer [18], Twins spatial-attention design [19], token-based Visual Transformers [21], comparative reviews [22], privacy-preserving ViT applications [24], and ViT image-classification case studies [23]—further demonstrate the breadth of transformer-based methods.

\textbf{Optimisation and activation functions.} Learning dynamics are also shaped by the non-linearity used in the hidden layers. Large empirical benchmarks [7, 8] and theoretical analyses [9, 10] highlight the strengths and weaknesses of ReLU, Swish, Softplus and related functions. Complementary studies of Softplus units [25], activation-function comparisons on MNIST [26], empirical analyses across vision tasks [27], and broader surveys [28] provide additional context on activation choice.

@@

\section{Methodology}
\subsection{Research Questions}
This study addresses the following key questions:
\begin{enumerate}
\item Does the choice between single-neuron and dual-neuron output configurations affect model accuracy, precision, recall, and F1 score in binary classification tasks?
\item Are there differences in convergence speed, stability, or learning dynamics between the two approaches?
\item Does one approach generalize better to unseen data than the other?
\item Are there specific types of binary classification problems or neural architectures where one approach consistently outperforms the other?
\end{enumerate}

@@

Pre-processing steps
\begin{itemize}
\item \textbf{Resize} All images are resized to the model‐specific input resolution (32×32 for Small CNN, 224×224 for ViT and ResNet-50).
\item \textbf{Normalisation} \texttt{mean=[0.485,0.456,0.406]}, \texttt{std=[0.229,0.224,0.225]} (ImageNet statistics).
\item \textbf{Augmentation (train only)} Random horizontal flip, rotation (±15°), colour jitter, and random affine translation (≤10 \%).
\end{itemize}

>>>>

% Graphics & tables
\usepackage{graphicx}
\usepackage{float} % Required for [H] figure placement
\usepackage{placeins} % For \FloatBarrier command
% \usepackage{booktabs} % Removed for compatibility

@@

\begin{abstract}
Binary classification is a classic problem in machine learning: given an input, the model predicts one of two outcomes. In binary classification, neural networks have traditionally implemented a single-output neuron with sigmoid activation. However, an equally valid alternative is to assign two output neurons with a softmax activation function. This study extensively compares the two architectural options with respect to their effect on the performance of models, their training dynamics, and generalization capabilities. From extensive experimentation on benchmark datasets across diverse neural network architectures, we probe under what conditions each approach might find favor. These observations add to knowledge with respect to design considerations for neural networks performing binary classification tasks.
\end{abstract}

\section{Introduction}
It is very common in the context of neural networks to handle binary classification problems through a single output neuron employing sigmoid activation; outputs nearing 0 are denoting a particular class while outputs approaching 1 are considered of the opposite type. Such a convention can definitely be abandoned; the second way to go about it consists of employing two output neurons under softmax activation, clearly modeling both the positive and "background" classes. Theoretically, both the configurations can solve the same problems, but it is the performance characteristics, training dynamics, and generalization capabilities using the different approaches that could possibly be different.

In this research, we try to systematically explain the difference between the two approaches and their influence on other qualities of model performance. We investigate whether the greater expressiveness afforded by the two-neuron output layer is truly an advantage or whether the single-neuron approach, with less expressiveness, presents some benefits of regularization and generalization.

@@

\subsection{Neural Network Output Layers for Classification}
Binary classification in neural networks has been historically implemented using one single output neuron with a  sigmoid activation [4]. The sigmoid function maps  output value of the networkbetween 0 and 1, which can be considered as the probability of the positive class. This approach is mathematically accurate and directly tied to logistic regression.

The alternative another approach uses two output neurons with softmax activation, which produces a probability distribution over the two classes [6]. While this is the standard approach for multi-class classification problems, its application to binary problems is less common in neural network architectures but theoretically possible.

Variants of these two-neuron softmax design have also been explored for binary problems. Memisevic et al. [6] introduced gated softmax in their work while Tyagi et al. [5] revisited the sigmoid–versus-softmax debate and showed that a properly tuned sigmoid–MSE objective can outperform softmax–cross-entropy. Further more investigations include Yang et al. [13], Klimo et al. [14], Maharjan et al. [15], and Hu et al. [16], which explore alternative loss functions, error-detecting output codes, and hardware-efficient softmax implementations.

\subsection{Related Research}
Prior research on binary image classification can be grouped into three types: \textbf{output-layer configurations}, \textbf{Vision Transformer (ViT) models}, and \textbf{optimisation \& activation-function studies}.

@@

\textbf{Vision Transformers.} Dosovitskiy et al.'s seminal ViT work [1] demonstrated that pure transformer architectures can match or surpass CNNs on large-scale image classification. Comprehensive surveys [2, 3] and data-efficient variants such as DeiT [11] together with the study by Khalil et al. [12] confirm ViT's competitiveness, motivating its inclusion in our comparison. The transformer paradigm itself originated with Vaswani et al. [20]. Additional surveys and architectural improvements—Khan et al. [17], the hierarchical Swin Transformer [18], Twins spatial-attention design [19], token-based Visual Transformers [21], comparative reviews [22], privacy-preserving ViT applications [24], and ViT image-classification case studies [23] further demonstrate the breadth of transformer-based methods for image classification.

\textbf{Optimisation and activation functions.} Learning dynamics are also created by the non linearity used in the hidden layers. Large empirical benchmarks [7, 8] and theoretical analyses [9, 10] highlight the strengths and weaknesses of ReLU, Swish, Softplus and related functions. Interrelated studies of Softplus units [25], activation-function comparisons on MNIST [26], empirical analyses across vision tasks [27], and broader surveys [28] provide additional context on activation choice.

@@

\section{Methodology}
\subsection{Research Questions}
This study addresses the following key questions:
\begin{enumerate}
\item Does the choice between single-neuron and dual-neuron output configurations affect model accuracy, precision, recall, and F1 score in binary classification tasks?
\item Are there any differences in convergence speed, stability, or learning dynamics between single-neuron and dual-neuron models?
\item Does one approach (single-neuron or dual-neuron) generalize better to unseen data than the other?
\item Are there any specific types of binary classification problems or neural architectures where one approach consistently outperforms the other?
\end{enumerate}

@@

\textbf{Pre-processing steps}
\begin{itemize}
\item \textbf{Resize} All images are resized to the model‐specific input resolution (32×32 for Small CNN, 224×224 for ViT and ResNet-50).
\item \textbf{Normalisation} \texttt{mean=[0.485,0.456,0.406]}, \texttt{std=[0.229,0.224,0.225]} (ImageNet statistics).
\item \textbf{Augmentation (train only)} Random horizontal flip, rotation (±15°), colour jitter, and random affine translation (≤10 \%).
\end{itemize}
